---
title: "tt13"
format: html
---

```{r}
library(tidyverse)
```


```{r}
financing_schemes <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2026/2026-04-21/financing_schemes.csv')
health_spending <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2026/2026-04-21/health_spending.csv')
spending_purpose <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2026/2026-04-21/spending_purpose.csv')
```

```{r}
health_spending |> 
  filter(iso3_code %in% c("ESP", "MEX", "FRA", "USA", "HTI")) |> 
  filter(unit == "% of current health expenditure") |> 
  filter(year == 2012) |> 
  ggplot(aes(x = country_name, y = value, fill = indicator_code)) + 
  geom_bar(stat = "identity")
```
```{r}
health_spending |> 
  filter(iso3_code %in% c("ESP", "MEX", "FRA", "USA", "HTI")) |> 
  filter(unit == "% of current health expenditure") |> 
  filter(year == 2012) |> 
  ggplot(aes(x = country_name, y = value, fill = indicator_code)) + 
  geom_bar(stat = "identity")

health_spending |> 
  filter(iso3_code %in% c("CUB", "HTI", "PAK", "SOM","KEN", "ETH", "MMR", "SSD", "NPL", "ARG")) |> 
  filter(indicator_code == "ext_che") |> 
  ggplot(aes(x = year, y = value, colour = iso3_code)) + 
  geom_point()+
  geom_line()
	

```


## Data Dictionary

### `financing_schemes.csv`

|variable |class |description |
|:--------|:-----|:-----------|
|country_name |character |Country or territory name as used by WHO. |
|iso3_code |character |ISO 3166-1 alpha-3 country code. |
|year |integer |Year of observation. |
|indicator_code |character |GHED indicator code. Suffix indicates unit: <code>_che</code> = % of CHE, <code>_usd2023</code> = constant 2023 US$. <br><br>Indicators: <br><code>hf1</code> — Government schemes and compulsory contributory health care financing schemes. <br><code>hf2</code> — Voluntary health care payment schemes. <br><code>hf3</code> — Household out-of-pocket payments (OOPS). <br><code>hf4</code> — Rest of the world financing schemes (non-resident). <br><code>hfnec</code> — Unspecified financing schemes (n.e.c.). |
|financing_scheme |character |Health care financing scheme. |
|value |double |Indicator value in the unit specified. |
|unit |character |Unit of measurement: <code>% of current health expenditure</code> or <code>constant 2023 US$</code>. |

### `health_spending.csv`

|variable |class |description |
|:--------|:-----|:-----------|
|country_name |character |Country or territory name as used by WHO. |
|iso3_code |character |ISO 3166-1 alpha-3 country code. |
|year |integer |Year of observation. |
|indicator_code |character |GHED indicator code. Suffix indicates unit: <code>_che</code> = % of CHE, <code>_usd2023</code> = constant 2023 US$. <br><br>Indicators: <br><code>che</code> — Current health expenditure (total, only <code>_usd2023</code>). <br><code>gghed</code> — Domestic general government health expenditure (GGHE-D). <br><code>pvtd</code> — Domestic private health expenditure (PVT-D). <br><code>ext</code> — External health expenditure (EXT). |
|expenditure_type |character |Type of health expenditure. |
|value |double |Indicator value in the unit specified. |
|unit |character |Unit of measurement: <code>% of current health expenditure</code> or <code>constant 2023 US$</code>. |

### `spending_purpose.csv`

|variable |class |description |
|:--------|:-----|:-----------|
|country_name |character |Country or territory name as used by WHO. |
|iso3_code |character |ISO 3166-1 alpha-3 country code. |
|year |integer |Year of observation. |
|indicator_code |character |GHED indicator code. Suffix indicates unit: <code>_che</code> = % of CHE, <code>_usd2023</code> = constant 2023 US$. Data available from 2016 onwards. <br><br>Indicators: <br><code>hc1</code> — Curative care. <br><code>hc2</code> — Rehabilitative care. <br><code>hc3</code> — Long-term care (health). <br><code>hc4</code> — Ancillary services. <br><code>hc5</code> — Medical goods. <br><code>hc6</code> — Preventive care. <br><code>hc7</code> — Governance and health system administration. <br><code>hc9</code> — Other health care services (n.e.c.). |
|spending_purpose |character |Health care function or purpose of spending. |
|value |double |Indicator value in the unit specified. |
|unit |character |Unit of measurement: <code>% of current health expenditure</code> or <code>constant 2023 US$</code>. |