---
title: "UK Met Weather Data"
author: "Federica & Prof. Jo Hardin"
format: 
  html:
     code-fold: true
execute: 
  warning: false
   
---


```{r}
#| echo: false
library(tidyverse)
library(showtext)
library(patchwork)
library(cowplot)
library(magick)
font_add_google("Playfair Display", "playfair")
showtext_auto()
```

### Data Wrangling
Loading the data
```{r}
historic_station_met <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2025/2025-10-21/historic_station_met.csv')
station_meta <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2025/2025-10-21/station_meta.csv')
```
This section defines the number of days in each month and creates month_breaks, a vector marking where each month ends. These breaks are later used as x-axis tick marks for labeling the polar plot by seasons or months.
```{r}
month_length <- c(31, 28, 31, 30, 31, 30,
                  31, 31, 30, 31, 30, 31)

month_breaks <- cumsum(month_length) - 30
```

### About this Week!

This week the dataset included data recorded from different UK weather stations starting in 1853 up until 2024. More details about this data set can be found below.

We decided to filter to observations after 2000s in order to enhance interpretability/visibility of the spirals.

### Plotting Logic

Rain plot
```{r}
rain <- historic_station_met |> 
  filter(station == "heathrow") |> 
  filter(year >= 2000) |> 
  mutate(date = ymd(paste(year, month, "01", sep = "-")),
         day_of_year = yday(date),
         # Handle December specially - extend to end of year instead of next year
         ymd_month = if_else(month == 12, 
                             ymd(paste(year, "12", "31", sep = "-")),
                             date %m+% months(1)),
         day_month = yday(ymd_month)) |> 
  ggplot() +
  geom_segment(aes(x = day_of_year, xend = day_month, 
                   y = date, yend = ymd_month, color = rain), linewidth = 2) +
  scale_color_distiller(palette = "BuPu",
                        direction = 1, 
                        name = "Rainfall (mm)") + 
  geom_label(data = . %>% filter(month == 1, year %% 5 == 0) ,
          aes(x = day_of_year, y = date, label = paste0(as.character(year), "→")),
          hjust = 1, size = 2.5, fontface = "bold",
          alpha = 0.5, fill = "#F5F7FA",
          label.size = NA) + 
  coord_polar() +
  theme_void() +
  scale_x_continuous(minor_breaks = month_breaks, 
                     breaks = month_breaks[c(3, 6, 9, 12)],
                     labels = c("Spring", "Summer", "Autumn", "Winter")) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1)))+
  theme(
    plot.background = element_rect(color = NA, fill = "#F5F7FA"),
    axis.text.x = element_text(color = "#2C3E50", size = 9, hjust = 10, 
                                family = "playfair", face = "bold"),
    legend.title = element_text(family = "playfair", face = "bold", 
                                size = 11, color = "#2C3E50"),
    legend.text = element_text(family = "playfair", size = 10, 
                               color = "#2C3E50"),
    plot.margin = margin(20, 30, 20, 90)
  )
```




Sun plot
```{r}
sun <- historic_station_met |> 
  filter(station == "heathrow") |> 
  filter(year >= 2000) |> 
  mutate(date = ymd(paste(year, month, "01", sep = "-")),
         day_of_year = yday(date),
         # Handle December specially - extend to end of year instead of next year
         ymd_month = if_else(month == 12, 
                             ymd(paste(year, "12", "31", sep = "-")),
                             date %m+% months(1)),
         day_month = yday(ymd_month)) |> 
  ggplot() +
  geom_segment(aes(x = day_of_year, xend = day_month, 
                   y = date, yend = ymd_month, color = sun), linewidth = 2) +
  geom_label(data = . %>% filter(month == 1, year %% 5 == 0) ,
          aes(x = day_of_year, y = date, label = paste0(as.character(year), "→")),
          hjust = 1, size = 2.5, fontface = "bold",
          alpha = 0.5, fill = "#F5F7FA",
          label.size = NA) +
  scale_color_distiller(palette = "YlOrRd", direction = 1, name = "Hours of Sunlight") + 
  coord_polar() +
  theme_void() +
  scale_x_continuous(minor_breaks = month_breaks, 
                     breaks = month_breaks[c(3, 6, 9, 12)],
                     labels = c("March", "June", "Sep.", "Dec.")) +
  theme(
    plot.background = element_rect(color = NA, fill = "#F5F7FA"),
    axis.text.x = element_text(color = "#2C3E50", size = 9, hjust = 10, 
                                family = "playfair", face = "bold"),
    legend.title = element_text(family = "playfair", face = "bold", 
                                size = 11, color = "#2C3E50"),
    legend.text = element_text(family = "playfair", size = 10, 
                               color = "#2C3E50"),
    plot.margin = margin(20, 30, 20, 90))
```


Combining plots
```{r}
rain_plot <- rain +
  theme(
    legend.position = "bottom",
    legend.direction = "horizontal",
    plot.margin = margin(20, 20, 20, 20)  
  )

sun_plot <- sun + 
  theme(
    legend.position = "bottom",
    legend.direction = "horizontal",
    plot.margin = margin(20, 20, 20, 20) 
  )

final_plot <- rain_plot + sun_plot + 
  plot_annotation(
    title = "London Weather\n Rain and Sunshine Patterns (2000-2024)",
    subtitle = "Monthly rainfall (left) and hours of sun (right) at Heathrow Weather Station\nRead each plot clockwise from the centre spiralling outward.",
    theme = theme(
      plot.background = element_rect(fill = "#F5F7FA", color = NA),
      plot.title = element_text(family = "playfair", face = "bold", 
                                size = 19, color = "#2C3E50", 
                                hjust = 0.5, margin = margin(t = 10, b = 5)),
      plot.subtitle = element_text(family = "playfair", size = 11, 
                                   color = "#5D6D7E", hjust = 0.5,
                                   margin = margin(b = 10)),
      plot.margin = margin(10, 30, 10, 30)  
    )
  )
```

Raindrops!
```{r}
rain_image <- image_read("raindrops.png") |> 
  image_fx("A*30") #reduces opacity

ggdraw(final_plot) +
  draw_image(
    rain_image,
    x = 0, y = 0,
    hjust = , halign = 1,
    vjust = 0, valign = 0,
    width = 1,
    
  )
```
This visualization juxtaposes monthly rainfall (left) and hours of sunlight (right) recorded at Heathrow, revealing seasonal and long-term weather patterns near London.

In this case, each spiral starts at the center in the year 2000 and extends outward to 2024. Moving clockwise, months progress cyclically around the circle. Colour encodes magnitude with darke blues meaning heavier rainfall and warmer reds meaning more sunlight. Labels mark key years every 5 years so that we can better keep track of the temporal progression. 

### What we see ☀️

* Rainfall peaks consistently during the winter months (November–February), while the sunniest months occur in late spring and summer (May–August).
* The spirals reveal recurring annual rhythms - London’s weather remains relatively stable in its seasonality over the 25-year period.
* Sunlight hours appear somewhat more variable, with several brighter summers after 2010, possibly reflecting warmer and drier seasons in the last decade.
* There is no strong long-term drying or wetting trend in rainfall magnitude, though some recent years (post-2018) show slightly drier winters compared to early 2000s.

## About this Data

This data was found on the [TidyTuesday Repository](https://github.com/rfordatascience/tidytuesday/blob/main/data/2025/2025-10-21/readme.md?plain=1)

The UK Met Office is the United Kingdom’s national weather and climate service,
providing forecasts, severe weather warnings, and climate science expertise. It
helps people, businesses, and governments make informed decisions to stay safe
and plan for the future. It was first established in 1854, making it one of the
oldest weather services in the world.

Data has been scraped straight from the Met Office website and cleaned in a
basic way. The few flags for "estimated" data and changing sunlight monitoring
techniques have been removed for simplicity, but are available from the raw
data. Also available is simple site metadata which includes the opening date and
[latitude and longitude](https://epsg.io/4326) of each station.

> Monthly Historical information for 37 UK Meteorological Stations. Most go back
to the early 1900s, but some go back as far as 1853. Station data files are
updated on a rolling monthly basis, around 10 days after the end of the month.
No allowances have been made for small site changes and developments in
instrumentation.


## Data Dictionary

### `historic_station_met.csv`

|variable |class     |description                           |
|:--------|:---------|:-------------------------------------|
|station  |character |Station ID; for binding with metadata.|
|year     |integer   |Year. |
|month    |integer   |Numeric month (1-12). |
|tmax     |double    |Mean daily maximum temperature (Degrees C). |
|tmin     |double    |Mean daily minimum temperature (Degrees C). |
|af       |double    |Days of air frost. |
|rain     |double    |Total rainfall (mm). |
|sun      |double    |Total sunshine duration (hours). |

### `station_meta.csv`

|variable     |class     |description                           |
|:------------|:---------|:-------------------------------------|
|station      |character |Station ID; for binding with measurement data. |
|station_name |character |Station name. |
|opened       |integer   |Year in which station opened. |
|lng          |double    |Decimal longitude. |
|lat          |double    |Decimal latitude. |

