---
title: "UK Met Weather Data"
author: "Federica & Prof. Jo Hardin"
format: 
  html:
     code-fold: true
execute: 
  warning: false
   
---
```{r}
library(tidyverse)
library(showtext)
library(patchwork)
library(cowplot)
library(magick)
font_add_google("Playfair Display", "playfair")
showtext_auto()
```


```{r}
#| echo: false

historic_station_met <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2025/2025-10-21/historic_station_met.csv')
station_meta <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2025/2025-10-21/station_meta.csv')
```
```{r}
month_length <- c(31, 28, 31, 30, 31, 30,
                  31, 31, 30, 31, 30, 31)

month_breaks <- cumsum(month_length) - 30
```


```{r}
rain <- historic_station_met |> 
  filter(station == "heathrow") |> 
  filter(year >= 2000) |> 
  mutate(date = ymd(paste(year, month, "01", sep = "-")),
         day_of_year = yday(date),
         # Handle December specially - extend to end of year instead of next year
         ymd_month = if_else(month == 12, 
                             ymd(paste(year, "12", "31", sep = "-")),
                             date %m+% months(1)),
         day_month = yday(ymd_month)) |> 
  ggplot() +
  geom_segment(aes(x = day_of_year, xend = day_month, 
                   y = date, yend = ymd_month, color = rain), linewidth = 2) +
  scale_color_distiller(palette = "BuPu",
                        direction = 1, 
                        name = "Rainfall (mm)") + 
  geom_label(data = . %>% filter(month == 1, year %% 5 == 0) ,
          aes(x = day_of_year, y = date, label = paste0(as.character(year), "→")),
          hjust = 1, size = 2.5, fontface = "bold",
          alpha = 0.5, fill = "#F5F7FA",
          label.size = NA) + 
  coord_polar() +
  theme_void() +
  scale_x_continuous(minor_breaks = month_breaks, 
                     breaks = month_breaks[c(3, 6, 9, 12)],
                     labels = c("Spring", "Summer", "Autumn", "Winter")) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1)))+
  theme(
    plot.background = element_rect(color = NA, fill = "#F5F7FA"),
    axis.text.x = element_text(color = "#2C3E50", size = 9, hjust = 10, 
                                family = "playfair", face = "bold"),
    legend.title = element_text(family = "playfair", face = "bold", 
                                size = 11, color = "#2C3E50"),
    legend.text = element_text(family = "playfair", size = 10, 
                               color = "#2C3E50"),
    plot.margin = margin(20, 30, 20, 90)
  )
```





```{r}
sun <- historic_station_met |> 
  filter(station == "heathrow") |> 
  filter(year >= 2000) |> 
  mutate(date = ymd(paste(year, month, "01", sep = "-")),
         day_of_year = yday(date),
         # Handle December specially - extend to end of year instead of next year
         ymd_month = if_else(month == 12, 
                             ymd(paste(year, "12", "31", sep = "-")),
                             date %m+% months(1)),
         day_month = yday(ymd_month)) |> 
  ggplot() +
  geom_segment(aes(x = day_of_year, xend = day_month, 
                   y = date, yend = ymd_month, color = sun), linewidth = 2) +
  geom_label(data = . %>% filter(month == 1, year %% 5 == 0) ,
          aes(x = day_of_year, y = date, label = paste0(as.character(year), "→")),
          hjust = 1, size = 2.5, fontface = "bold",
          alpha = 0.5, fill = "#F5F7FA",
          label.size = NA) +
  scale_color_distiller(palette = "YlOrRd", direction = 1, name = "Hours of Sunlight") + 
  coord_polar() +
  theme_void() +
  scale_x_continuous(minor_breaks = month_breaks, 
                     breaks = month_breaks[c(3, 6, 9, 12)],
                     labels = c("March", "June", "Sep.", "Dec.")) +
  theme(
    plot.background = element_rect(color = NA, fill = "#F5F7FA"),
    axis.text.x = element_text(color = "#2C3E50", size = 9, hjust = 10, 
                                family = "playfair", face = "bold"),
    legend.title = element_text(family = "playfair", face = "bold", 
                                size = 11, color = "#2C3E50"),
    legend.text = element_text(family = "playfair", size = 10, 
                               color = "#2C3E50"),
    plot.margin = margin(20, 30, 20, 90))
```

```{r}
rain_plot <- rain +
  theme(
    legend.position = "bottom",
    legend.direction = "horizontal",
    plot.margin = margin(20, 20, 20, 20)  
  )

sun_plot <- sun + 
  theme(
    legend.position = "bottom",
    legend.direction = "horizontal",
    plot.margin = margin(20, 20, 20, 20) 
  )

final_plot <- rain_plot + sun_plot + 
  plot_annotation(
    title = "London Weather\n Rain and Sunshine Patterns (2000-2024)",
    subtitle = "Monthly rainfall (left) and hours of sun (right) at Heathrow Weather Station\nRead each plot clockwise from the centre spiralling outward.",
    theme = theme(
      plot.background = element_rect(fill = "#F5F7FA", color = NA),
      plot.title = element_text(family = "playfair", face = "bold", 
                                size = 19, color = "#2C3E50", 
                                hjust = 0.5, margin = margin(t = 10, b = 5)),
      plot.subtitle = element_text(family = "playfair", size = 11, 
                                   color = "#5D6D7E", hjust = 0.5,
                                   margin = margin(b = 10)),
      plot.margin = margin(10, 30, 10, 30)  
    )
  )
```

```{r}
rain_image <- image_read("raindrops.png") |> 
  image_fx("A*30") #reduces opacity

ggdraw(final_plot) +
  draw_image(
    rain_image,
    x = 0, y = 0,
    hjust = , halign = 1,
    vjust = 0, valign = 0,
    width = 1,
    
  )
```





### About this Data

This data was found on the [TidyTuesday Repository](https://github.com/rfordatascience/tidytuesday/blob/main/data/2025/2025-10-21/readme.md?plain=1)

The UK Met Office is the United Kingdom’s national weather and climate service,
providing forecasts, severe weather warnings, and climate science expertise. It
helps people, businesses, and governments make informed decisions to stay safe
and plan for the future. It was first established in 1854, making it one of the
oldest weather services in the world.

Data has been scraped straight from the Met Office website and cleaned in a
basic way. The few flags for "estimated" data and changing sunlight monitoring
techniques have been removed for simplicity, but are available from the raw
data. Also available is simple site metadata which includes the opening date and
[latitude and longitude](https://epsg.io/4326) of each station.

> Monthly Historical information for 37 UK Meteorological Stations. Most go back
to the early 1900s, but some go back as far as 1853. Station data files are
updated on a rolling monthly basis, around 10 days after the end of the month.
No allowances have been made for small site changes and developments in
instrumentation.

### `historic_station_met.csv`

|variable |class     |description                           |
|:--------|:---------|:-------------------------------------|
|station  |character |Station ID; for binding with metadata.|
|year     |integer   |Year. |
|month    |integer   |Numeric month (1-12). |
|tmax     |double    |Mean daily maximum temperature (Degrees C). |
|tmin     |double    |Mean daily minimum temperature (Degrees C). |
|af       |double    |Days of air frost. |
|rain     |double    |Total rainfall (mm). |
|sun      |double    |Total sunshine duration (hours). |

### `station_meta.csv`

|variable     |class     |description                           |
|:------------|:---------|:-------------------------------------|
|station      |character |Station ID; for binding with measurement data. |
|station_name |character |Station name. |
|opened       |integer   |Year in which station opened. |
|lng          |double    |Decimal longitude. |
|lat          |double    |Decimal latitude. |