---
title: "Selected British Literary Prizes (1990-2022)"
author: "Federica & TidyTuesday Friends"
date: 10/28/2025
date-format: "D MMM, YYYY"
format: 
  html:
     code-fold: true
execute: 
  warning: false
---


```{r}
#| echo: FALSE
library(tidyverse)
library(ggalluvial)
library(ggsankey)
library(showtext)
font_add_google("Playfair Display", "playfair")
showtext_auto()
```
### Data Wrangling
Here we load the prizes dataset from TidyTuesday and clean up missing degree entries. Then, we prefix degree names with letters (a–i) so we can later enforce a custom order - otherwise, `ggsankey` reorders them alphabetically.
```{r}
prizes <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2025/2025-10-28/prizes.csv') |> 
  mutate(highest_degree = ifelse(is.na(highest_degree), "unknown", 
                                 highest_degree)) |> 
  mutate(highest_degree = forcats::fct_recode(highest_degree,  
                                 "a none" = "none",
                                 "b unknown" = "unknown", 
                                 "c Diploma" = "Diploma", 
                                 "d Certificate of Education" = "Certificate of Education", 
                                 "e Bachelors" = "Bachelors", 
                                 "f Masters" = "Masters", 
                                 "g Juris Doctor" = "Juris Doctor", 
                                 "h MD" = "MD", 
                                 "i Doctorate" = "Doctorate")) 
  
```

Here we isolate Booker Prize data only and reshape it with make_long() to a "long" format suitable for `ggsankey`.
Each column (gender, ethnicity, education, field) becomes a stage in the flow diagram.
```{r}
booker <-  prizes |> 
  filter(prize_name == "Booker Prize", person_role == "winner") |> 
  rename(
    Gender = "gender",
    `Ethnicity Macro` = "ethnicity_macro",
    `Highest Degree` = "highest_degree",
    `Degree Field` = "degree_field"
  ) |> 
  ggsankey::make_long(Gender, `Ethnicity Macro`, `Highest Degree`,  `Degree Field`) 
```

### Plotting Logic
```{r}
booker |> 
  ggplot(aes(x = x,
             next_x = next_x, 
             node = node, 
             next_node = next_node, 
             fill = factor(node), 
             label = node)) + 
  geom_sankey(flow.alpha = .3,
            show.legend = FALSE) +
  geom_sankey_label(show.legend = FALSE,
                    aes(
                      hjust = case_when(
                        x == "gender" ~.7,
                        x == "degree_field" ~0,
                        TRUE ~ .5
                      )
                    ),
                    size = 3,
                  family = "playfair",   # serif, literary feel
                  colour = "gray15",
                  fill = alpha("cornsilk", 0.9)
                    ) +
       scale_fill_viridis_d(option = "C") +
       scale_x_discrete(expand = expansion(mult = c(0.15, 0.21))) +
       theme_sankey(base_size = 12) +
       theme(legend.position = "none",
             axis.title = element_blank(),
             axis.text.y = element_blank(),
             axis.ticks = element_blank(),
             panel.grid = element_blank(),
             plot.background = element_rect(fill = "#f8f5ef", color = NA), 
             panel.background = element_rect(fill = "#f8f5ef", color = NA),
             plot.title = element_text(
               family = "playfair",
               face = "bold",
               size = 16,
               hjust = 0.5,
               color = "#3a2e1f"   # deep brown like old ink
              ),
            plot.subtitle = element_text(
              family = "playfair",
              hjust = 0.5,
              size = 11,
              color = "#5c4a34"
            ),
            ) +
      labs(title = "A Look at Booker Prize Award Winners 1990 - 2022",
           subtitle = "Mapping gender, ethnicity, and education backgrounds across decades")
```

This Sankey diagram explores how gender, ethnicity, and educational background intersect among Booker Prize award winners between 1990 and 2022. Each flow traces the path from an author’s gender to their ethnicity, highest degree, and finally, their field of study - revealing how academic and cultural backgrounds have shaped the landscape of British literary recognition.

The visualisation was built in R using `ggsankey` and customised extensively for a bookish, parchment-like theme. The categorical ordering of degrees (e.g., “a none,” “b unknown,” “e Bachelors,” etc.) was manually imposed to reflect an educational hierarchy, we weren't able to figure out how to override the library's settings when it came to within category ordering.

The design draws inspiration from [Bioinfo Scrounger’s Alluvial & Sankey Guide](https://www.bioinfo-scrounger.com/archives/alluvial-and-Sankey/), which helped shape the coding approach and aesthetic structure of this plot.

### What we see 📖
* Most Booker Prize winners hold higher education degrees, especially Masters and Bachelors degrees.
* A large proportion of winners identify as White British or Non-UK White, reflecting ongoing questions of representation in literary prestige.
* There’s still a noticeable gender imbalance, with men representing the vast majority of awardees.


## About the Data

This data was found on the [TidyTuesday Repository](https://github.com/rfordatascience/tidytuesday/blob/main/data/2025/2025-10-28/readme.md?plain=1)

This week we are exploring data related to the Selected British Literary Prizes (1990-2022) dataset which comes from the [Post45 Data Collective](https://data.post45.org/posts/british-literary-prizes/). 

> "This dataset contains primary categories of information on individual authors comprising gender, sexuality, UK residency, ethnicity, geography and details of educational background, including institutions where the authors acquired their degrees and their fields of study. Along with other similar projects, we aim to provide information to assess the cultural, social and political factors determining literary prestige. Our goal is to contribute to greater transparency in discussions around diversity and equity in literary prize cultures." 

## Data Dictionary

### `prizes.csv`

|variable              |class     |description                           |
|:---------------------|:---------|:-------------------------------------|
|prize_id              |integer    |Unique prize identifier used in the SBLP dataset. |
|prize_alias           |character |Name of the prize awarded, regularized to the most current name. |
|prize_name            |character |Name of the prize awarded, at the time of award. |
|prize_institution     |character |Institution that sponsored the prize. |
|prize_year            |integer  |Year the prize was awarded. |
|prize_genre           |character |Genre category of book that the prize was awarded to. |
|person_id             |character |Unique author identifier used in the SBLP dataset, assigned in order of entity entry to the dataset. |
|person_role           |character |Whether author was shortlisted or won the prize. |
|last_name             |character |Family name of author. |
|first_name            |character |Given name of author. |
|name                  |character |Full name author in family name, given name format. |
|gender                |character |Author’s gender, as self-declared/publicly available. |
|sexuality             |character |Author’s sexuality, as self-declared/publicly available. |
|uk_residence          |logical   |Whether the author holds residence status in the UK at the time of data gathering. |
|ethnicity_macro       |character |Ethnicity macro category, as created for this dataset. |
|ethnicity             |character |Ethnicity as self-declared/publicly available. |
|highest_degree        |character |Highest level of post-secondary education. |
|degree_institution    |character |Institution from which the highest degree was attained. |
|degree_field_category |character |Degree macro category, as created for this dataset. |
|degree_field          |character |Field of study, as self-declared/publicly available. |
|viaf                  |character |Virtual internet authority file code. |
|book_id               |character |Unique book identifier used in the SBLP dataset. |
|book_title            |character |Title of the awarded or shortlisted book. |
