---
title: "tt9"
format: html
---

```{r}
library(tidyverse)
library(rvest)
agriculture <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2026/2026-02-17/dataset.csv') |> 
              rename(year = year_ended_june)
```
```{r}
agriculture |> 
  filter(str_detect(value_label, "(?i)^number")) |> 
  group_by(measure) |> 
  summarise(count_distinct = n()) |> 
  arrange(desc(count_distinct)) |> 
  slice_head(n =10) |> 
  pull(measure) 
```


```{r}
pop_nz <- read_html("https://commons.wikimedia.org/wiki/Data:New_Zealand_Total_Population.tab") |>
  html_elements("table") |> 
  html_table() |> 
  as.data.frame() |> 
  slice(-c(1:2)) |> 
  mutate(year = as.numeric(x),
         population = as.numeric(y1) * 1000000) |> 
  select(year, population)
```

```{r}
agric_pop <- agriculture |> 
  left_join(pop_nz, by = join_by(year)) |> 
   filter(measure %in% c("Total Lambs Marked and/or Tailed", "Total Sheep"                            
, "Breeding Ewes 2 Tooth and Over put to Ram"        
, "Number of Farm Holdings"                          
, "Rising 1 Year Old Dairy Heifers and Heifer Calves"
, "Beef Cows and Heifers, Bred from"                 
, "Total Beef Cattle"                                
, "Total Dairy Cattle (including Bobby Calves)"      
, "Total Pigs"                                       
, "Dairy Cows and Heifers, in Milk or Calf"))
```


```{r}
agric_pop |> 
  ggplot( ) + 
  geom_line(aes(x = year, y = value, color = measure)) + 
  geom_line(aes(x = year, y = population))
```




# Agricultural Production Statistics in New Zealand

This week we are exploring [agriculture production statistics in New Zealand](https://figure.nz/table/TSQ8lkuKnyzfERF3) using data compiled from [StatsNZ](https://datainfoplus.stats.govt.nz/item/nz.govt.stats/eb7208cf-2f0f-416a-bd77-2fdbb9043255?_ga=2.264735308.2106648024.1715287902-1399521469.1678132138). 

Sheep have long outnumbered people in New Zealand, but the [ratio of sheep to people peaked in the 1980s and has been in steady decline](https://www.rnz.co.nz/news/country/560252/gap-between-people-and-sheep-rapidly-closing)

> The gap between people and sheep in New Zealand is rapidly closing. There's now about 4.5 sheep to every person in New Zealand compared to a peak of 22 sheep per person in the 1980s, that's according to figures released by Stats NZ this week.

- Is sheep production unique in its decline? Do other types of meat production show the same pattern?
- Which agricultural industries have shown the most production growth?

Thank you to [Jen Richmond](https://github.com/jenrichmond) for curating this week's dataset.


## Data Dictionary

### `dataset.csv`

|variable        |class     |description                           |
|:---------------|:---------|:-------------------------------------|
|year_ended_june |integer    |Year |
|measure         |character |Agricultural production category |
|value           |integer    |Number of units produced |
|value_label     |character |Unit of production |