---
title: "Lead concentration in Flint water samples in 2015"
author: "Federica & Prof. Hardin"
date: 11/04/2025
date-format: "D MMM, YYYY"
format: 
  closeread-html:
      theme: 
      - default
      - closeread-styles.scss
execute: 
  warning: false
---


```{r}
library(tidyverse)
library(gganimate)
library(emojifont)
load.fontawesome()
flint_mdeq <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2025/2025-11-04/flint_mdeq.csv')
```

```{r}
#| echo: false

# Calculate quantiles
lead_90_all <- quantile(flint_mdeq$lead, 0.9, na.rm = TRUE)
lead_90_filtered <- quantile(flint_mdeq$lead2, 0.9, na.rm = TRUE)

# Create base data for plots
plot_data <- flint_mdeq |>
  arrange(sample) |>
  mutate(lead_bin = floor(lead / 5) * 5) |>
  group_by(lead_bin) |>
  mutate(cumulative_count = row_number()) |>
  ungroup() |>
  mutate(
    removed = is.na(lead2),  
    lead_level = lead
  )
```

In 2015, water samples were collected from homes across Flint, Michigan to test for lead contamination.

:::{.cr-section}
```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 6

# Create animation data
animation_data <- plot_data |>
  mutate(
    start_x = 0,  
    start_y = 50,   
    end_x = lead_bin,
    end_y = cumulative_count
  ) |>
  rowwise() |>
  reframe(
    sample = sample,
    time = c(sample - 1, sample),  
    x = c(start_x, end_x),
    y = c(start_y, end_y),
    lead_level = lead_level
  ) |>
  ungroup()

p1 <- animation_data |>
  ggplot(aes(x = x, y = y, group = sample, color = lead_level)) +
  geom_text(aes(label = fontawesome('fa-tint')),  
            family = 'fontawesome-webfont',
            size = 8, 
            alpha = 0.8) +
  scale_color_gradient2(
    low = "steelblue",      
    mid = "orange",         
    high = "darkorange",    
    midpoint = 15,          
    limits = c(0, max(flint_mdeq$lead, na.rm = TRUE)),
    name = "Lead (ppb)"
  ) +
  transition_time(time) +  
  shadow_wake(wake_length = 2, alpha = 0.4) +  
  shadow_mark(past = TRUE, alpha = 0.8, size = 8) +  
  ease_aes('cubic-out') +  
  labs(x = "Lead (ppb)", 
       y = "Count",
       title = "Water Samples Collected Across Flint") +
  theme_minimal() + 
  theme(panel.background = element_rect(fill = "aliceblue"))

animate(p1, duration = 15, fps = 25, loop = FALSE)
```
:::



[The EPA action level for lead in drinking water is 15 parts per billion (ppb).]{.cr-focus}
:::{.cr-section}
```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 6

plot_data |>
  ggplot(aes(x = lead_bin, y = cumulative_count, color = lead_level)) +
  geom_vline(xintercept = 15, linetype = "dashed", color = "red", linewidth = 1.5) +
  annotate("text", x = 15, y = Inf, label = "EPA Action Level (15 ppb)", 
           vjust = 1.5, hjust = -0.1, color = "red", size = 5, fontface = "bold") +
  geom_text(aes(label = fontawesome('fa-tint')),  
            family = 'fontawesome-webfont',
            size = 8, 
            alpha = 0.8) +  
  scale_color_gradient2(
    low = "steelblue",
    mid = "lightblue",
    high = "chocolate2",   
    midpoint = 15,          
    limits = c(0, max(flint_mdeq$lead, na.rm = TRUE)),
    name = "Lead (ppb)"
  ) + 
  labs(x = "Lead (ppb)", 
       y = "Count",
       title = "Many Samples Exceeded Safe Levels") +
  theme_minimal() + 
  theme(panel.background = element_rect(fill = "aliceblue"))
```
When more than 10% of samples exceed this level, action is required.

:::

[The key metric is the 90th percentile of lead levels.]{.cr-focus}
For all samples, this was well above the action level.

:::{.cr-section}
```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 6

plot_data |>
  ggplot(aes(x = lead_bin, y = cumulative_count, color = lead_level)) +
  geom_vline(xintercept = 15, linetype = "dashed", color = "red", linewidth = 1) +
  geom_vline(xintercept = lead_90_all, linetype = "solid", color = "darkred", linewidth = 1.5) +
  annotate("text", x = lead_90_all, y = Inf, label = paste0("90th Percentile\n(", round(lead_90_all, 1), " ppb)"), 
           vjust = 1.5, hjust = -0.1, color = "darkred", size = 5, fontface = "bold") +
  geom_text(aes(label = fontawesome('fa-tint')),  
            family = 'fontawesome-webfont',
            size = 8, 
            alpha = 0.8) +  # Remove color from here
  scale_color_gradient2(
    low = "steelblue",
    mid = "lightblue",
    high = "chocolate2",   
    midpoint = 15,          
    limits = c(0, max(flint_mdeq$lead, na.rm = TRUE)),
    name = "Lead (ppb)"
  )  +
  labs(x = "Lead (ppb)", 
       y = "Count",
       title = "90th Percentile Shows Severe Contamination") +
  theme_minimal() + 
  theme(panel.background = element_rect(fill = "aliceblue"))
```
:::


[But some samples were excluded from the official analysis.]{.cr-focus} These high-lead samples were removed from the dataset.

:::{.cr-section}
```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 6

plot_data |>
  ggplot(aes(x = lead_bin, y = cumulative_count)) +
  geom_vline(xintercept = 15, linetype = "dashed", color = "red", linewidth = 1) +
  geom_vline(xintercept = lead_90_all, linetype = "solid", color = "darkred", linewidth = 1.5) +
  annotate("text", x = lead_90_all, y = Inf, label = paste0("90th Percentile\n(", round(lead_90_all, 1), " ppb)"), 
           vjust = 1.5, hjust = -0.1, color = "darkred", size = 5, fontface = "bold") +
  geom_text(data = filter(plot_data, removed), 
            aes(label = fontawesome('fa-tint')),
            color = "salmon",
            family = 'fontawesome-webfont',
            size = 10, 
            alpha = 1) +
  # Show kept samples faded
  geom_text(data = filter(plot_data, !removed), 
            aes(label = fontawesome('fa-tint'), color = lead_level),
            family = 'fontawesome-webfont',
            size = 8, 
            alpha = 0.3) + scale_color_gradient2(
    low = "steelblue",
    mid = "lightblue",
    high = "chocolate2",   
    midpoint = 15,          
    limits = c(0, max(flint_mdeq$lead, na.rm = TRUE)),
    name = "Lead (ppb)"
  )  +
  labs(x = "Lead (ppb)", 
       y = "Count",
       title = "Highlighted: Samples Excluded from Official Analysis",
       subtitle = "These high-contamination samples were removed") +
  theme_minimal() + 
  theme(panel.background = element_rect(fill = "aliceblue"))
```


:::

[With those samples removed, the 90th percentile dropped to 11.4 ppb]{.cr-focus} - conveniently just below the EPA action level.
::: {.cr-section}

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 6

plot_data |>
  filter(!removed) |>  # Only show kept samples
  ggplot(aes(x = lead_bin, y = cumulative_count, color = lead_level)) +
  geom_vline(xintercept = 15, linetype = "dashed", color = "red", linewidth = 1) +
  geom_vline(xintercept = lead_90_filtered, linetype = "solid", color = "darkgreen", linewidth = 1.5) +
  annotate("text", x = 15, y = Inf, label = "EPA Action Level", 
           vjust = 1.5, hjust = -0.1, color = "red", size = 4) +
  annotate("text", x = lead_90_filtered, y = Inf, label = paste0("New 90th Percentile\n(", round(lead_90_filtered, 1), " ppb)"), 
           vjust = 1.5, hjust = 1.1, color = "darkgreen", size = 5, fontface = "bold") +
  geom_text(aes(label = fontawesome('fa-tint')),  
            family = 'fontawesome-webfont',
            size = 8, 
            alpha = 0.8) +
  scale_color_gradient2(
    low = "steelblue",      
    mid = "orange",         
    high = "darkorange",    
    midpoint = 15,          
    limits = c(0, max(flint_mdeq$lead, na.rm = TRUE)),
    name = "Lead (ppb)"
  ) +
  labs(x = "Lead (ppb)", 
       y = "Count",
       title = "After Exclusions: Below Action Level",
       subtitle = "The 90th percentile now appears acceptable") +
  theme_minimal() + 
  theme(panel.background = element_rect(fill = "aliceblue"))
```
:::



## The data

```{r}
#|echo: FALSE
flint_mdeq <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2025/2025-11-04/flint_mdeq.csv')
flint_vt <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2025/2025-11-04/flint_vt.csv')
```



## TidyTuesday Data

This dataset can be found on the [tidytuesday repo](https://github.com/rfordatascience/tidytuesday/blob/main/data/2025/2025-11-04/readme.md).

This week we are exploring lead levels in water samples collected in Flint, Michigan in 2015. 
The data comes from a paper by [Loux and Gibson (2018)](https://onlinelibrary.wiley.com/doi/pdf/10.1111/test.12187?casa_token=av3lP7OmqS0AAAAA:QAF3yU5kGzsUkqi1VlXkMlIN8ExolHZBSkdJ3hIHnlptUES57dGVoXjE3qdwPPHtLHNRd9VvX1x8f6VN).

The data this week includes samples collected by the Michigan Department of Environment (MDEQ) and data from a citizen science project coordinated by Prof Marc Edwards and colleagues at Virginia Tech. 
Community-sourced samples were collected after concerns were raised about the MDEQ excluding samples from their data. You can read about the ["murky" story behind this data here](https://academic.oup.com/jrssig/article/14/2/16/7029247). 

## Data Dictionary

### `flint_mdeq.csv`

|variable |class     |description                           |
|:--------|:---------|:-------------------------------------|
|sample   |double |sample number |
|lead     |double |lead level in parts per billion (all samples) |
|lead2    |double |lead level in parts per billion (2 samples removed) |
|notes    |character |comment about data removal |

### `flint_vt.csv`

|variable |class   |description                           |
|:--------|:-------|:-------------------------------------|
|sample   |integer |sample number |
|lead     |double  |lead levels in parts per billion (ppb)|