---
title: "Lead concentration in Flint water samples in 2015"
author: "Federica & Prof. Hardin"
date: 11/04/2025
date-format: "D MMM, YYYY"
execute: 
  warning: false
---

```{r}
#| include: false
library(tidyverse)
library(emojifont)
load.fontawesome()

# Load data
flint_mdeq <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2025/2025-11-04/flint_mdeq.csv')

# Calculate values
lead_90_all <- quantile(flint_mdeq$lead, 0.9)
lead_90_filtered <- quantile(flint_mdeq$lead2, 0.9, na.rm = TRUE)

# Prepare plot data
plot_data <- flint_mdeq |>
  arrange(sample) |>
  mutate(lead_bin = floor(lead / 5) * 5) |>
  group_by(lead_bin) |>
  mutate(cumulative_count = row_number()) |>
  ungroup() |>
  mutate(
    removed = is.na(lead2), 
    lead_level = lead
  )

x_max <- max(plot_data$lead_bin, na.rm = TRUE)
y_max <- max(plot_data$cumulative_count, na.rm = TRUE)
```

## The Flint Water Crisis

@cr-section

In 2015, water samples were collected from homes across Flint, Michigan to test for lead contamination.
```{r}
#| label: fig-samples
#| cr-id: samples
#| echo: false
p1 <- plot_data |>
  ggplot(aes(x = lead_bin, y = cumulative_count, color = lead_level)) +
  geom_text(aes(label = fontawesome('fa-tint')),
            family = 'fontawesome-webfont',
            size = 8, 
            alpha = 0.8) +
  scale_color_gradient2(
    low = "steelblue",
    mid = "lightblue",
    high = "chocolate2",   
    midpoint = 15,          
    limits = c(0, max(flint_mdeq$lead, na.rm = TRUE)),
    name = "Lead (ppb)"
  ) +
  coord_cartesian(xlim = c(0, x_max), ylim = c(0, y_max)) +
  labs(x = "Lead (ppb)", y = "Count") +
  theme_minimal() + 
  theme(panel.background = element_rect(fill = "aliceblue"),
        legend.position = "none")
p1
```

[Each droplet represents one water sample from a Flint home.]{.cr-sticky}

[The EPA set an action level of 15 parts per billion (ppb) for lead.]{.cr-sticky}
```{r}
#| label: fig-epa
#| cr-id: epa
#| echo: false
p2 <- plot_data |>
  ggplot(aes(x = lead_bin, y = cumulative_count, color = lead_level)) +
  geom_vline(xintercept = 15, linetype = "dashed", color = "red", linewidth = 1.5) +
  annotate("text", x = 15, y = Inf, label = "EPA Action Level (15 ppb)", 
           vjust = 1.5, hjust = -0.1, color = "red", size = 5, fontface = "bold") +
  geom_text(aes(label = fontawesome('fa-tint')),
            family = 'fontawesome-webfont',
            size = 8, 
            alpha = 0.8) +
  scale_color_gradient2(
    low = "steelblue",
    mid = "lightblue",
    high = "chocolate2",   
    midpoint = 15,          
    limits = c(0, max(flint_mdeq$lead, na.rm = TRUE)),
    name = "Lead (ppb)"
  ) +
  coord_cartesian(xlim = c(0, x_max), ylim = c(0, y_max)) +
  labs(x = "Lead (ppb)", y = "Count") +
  theme_minimal() + 
  theme(panel.background = element_rect(fill = "aliceblue"),
        legend.position = "none")
p2
```

[Water systems must keep 90% of samples below this threshold.]{.cr-sticky}
```{r}
#| label: fig-90th
#| cr-id: percentile
#| echo: false
p3 <- plot_data |>
  ggplot(aes(x = lead_bin, y = cumulative_count, color = lead_level)) +
  geom_vline(xintercept = 15, linetype = "dashed", color = "red", linewidth = 1) +
  geom_vline(xintercept = lead_90_all, linetype = "solid", color = "darkred", linewidth = 1.5) +
  annotate("text", x = lead_90_all, y = Inf, 
           label = paste0("90th Percentile\n(", round(lead_90_all, 1), " ppb)"), 
           vjust = 1.5, hjust = -0.1, color = "darkred", size = 5, fontface = "bold") +
  geom_text(aes(label = fontawesome('fa-tint')),
            family = 'fontawesome-webfont',
            size = 8, 
            alpha = 0.8) +
  scale_color_gradient2(
    low = "steelblue",
    mid = "lightblue",
    high = "chocolate2",   
    midpoint = 15,          
    limits = c(0, max(flint_mdeq$lead, na.rm = TRUE)),
    name = "Lead (ppb)"
  ) +
  coord_cartesian(xlim = c(0, x_max), ylim = c(0, y_max)) +
  labs(x = "Lead (ppb)", y = "Count") +
  theme_minimal() + 
  theme(panel.background = element_rect(fill = "aliceblue"),
        legend.position = "none")
p3
```

[But there was a problem. Two samples with high lead levels were removed from the official data.]{.cr-sticky}
```{r}
#| label: fig-removed
#| cr-id: removed
#| echo: false

p4 <- plot_data |>
  ggplot(aes(x = lead_bin, y = cumulative_count)) +
  geom_vline(xintercept = 15, linetype = "dashed", color = "red", linewidth = 1) +
  geom_vline(xintercept = lead_90_all, linetype = "solid", color = "darkred", linewidth = 1.5) +
  annotate("text", x = lead_90_all, y = Inf, label = paste0("90th Percentile\n(", round(lead_90_all, 1), " ppb)"), 
           vjust = 1.5, hjust = -0.1, color = "darkred", size = 5, fontface = "bold") +
  # kept samples faded
  geom_text(data = filter(plot_data, !removed), 
            aes(label = fontawesome('fa-tint'), color = lead_level),
            family = 'fontawesome-webfont',
            size = 8, 
            alpha = 0.3) + scale_color_gradient2(
    low = "steelblue",
    mid = "lightblue",
    high = "chocolate2",   
    midpoint = 15,          
    limits = c(0, max(flint_mdeq$lead, na.rm = TRUE)),
    name = "Lead (ppb)"
  )  +
geom_text(data = filter(plot_data, removed), 
            aes(label = fontawesome('fa-tint')),
            color = "salmon",
            family = 'fontawesome-webfont',
            size = 10, 
            alpha = 1) +
  coord_cartesian(xlim = c(0, x_max), ylim = c(0, y_max)) +  
  labs(x = "Lead (ppb)", 
       y = "Count") +
  theme_minimal() + 
  theme(panel.background = element_rect(fill = "aliceblue"))

p4
```

[After removing these samples, the 90th percentile dropped significantly—making the water appear safer than it was.]{.cr-sticky}
```{r}
#| label: fig-new
#| cr-id: filtered
#| echo: false

p5 <- plot_data |>
  filter(!removed) |> 
  ggplot(aes(x = lead_bin, y = cumulative_count, color = lead_level)) +
  scale_color_gradient2(
    low = "steelblue",
    mid = "lightblue",
    high = "chocolate2",   
    midpoint = 15,          
    limits = c(0, max(flint_mdeq$lead, na.rm = TRUE)),
    name = "Lead (ppb)"
  )  + 
  geom_vline(xintercept = 15, linetype = "dashed", color = "red", linewidth = 1) +
  geom_vline(xintercept = lead_90_filtered, linetype = "solid", color = "darkgreen", linewidth = 1.5) +
  annotate("text", x = lead_90_filtered, y = Inf, label = paste0("New 90th Percentile\n(", round(lead_90_filtered, 1), " ppb)"), 
           vjust = 1.5, hjust = -.6, color = "darkgreen", size = 5, fontface = "bold") +
   geom_text(data = filter(plot_data, !removed), 
            aes(label = fontawesome('fa-tint'), color = lead_level),
            family = 'fontawesome-webfont',
            size = 8, 
            alpha = 0.8) + 
  coord_cartesian(xlim = c(0, x_max), ylim = c(0, y_max)) +  
  labs(x = "Lead (ppb)", 
       y = "Count") +
  theme_minimal() + 
  theme(panel.background = element_rect(fill = "aliceblue"))
p5
```

@cr-section

## The Impact

The official 90th percentile was reported as **`r round(lead_90_filtered, 1)` ppb**, just below the EPA action level. 

But with all samples included, it was actually **`r round(lead_90_all, 1)` ppb**—well above the safe threshold.

This data manipulation helped hide the severity of Flint's water crisis, leading to prolonged exposure of residents to dangerous lead levels.

---

### Data Source

Data from [Loux and Gibson (2018)](https://onlinelibrary.wiley.com/doi/pdf/10.1111/test.12187) via [TidyTuesday](https://github.com/rfordatascience/tidytuesday/blob/main/data/2025/2025-11-04/readme.md).