---
title: "tt6"
author: "Federica"
date: 11/11/2025
date-format: "D MMM, YYYY"
format: 
  html:
     code-fold: true
execute: 
  warning: false
---

```{r}
library(tidyverse)
library(plotly)
```


```{r}
tb <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2025/2025-11-11/who_tb_data.csv')
```

```{r}
clean_tb <- tb |> 
  filter(!is.na(e_inc_100k)) |> 
  mutate(year = as.integer(year))

global_totals <- clean_tb |> 
  group_by(year) |> 
  summarise(total_inc = sum(e_inc_num, na.rm = TRUE))
```

```{r}
map_plot <- plot_ly() |> 
  add_trace(
    data = clean_tb,
    type = "choropleth",
    locations = ~iso3,
    z = ~e_inc_100k,
    frame = ~year,
    text = ~paste0(
      country, "<br>",
      "Incidence: ", round(e_inc_100k, 1), " per 100k"
    ),
    hoverinfo = "text",
    colorscale = "Viridis",
    marker = list(line = list(width = 0))
  ) |> 
  layout(
    geo = list(
      projection = list(type = "natural earth"),
      showframe = FALSE,
      showcoastlines = FALSE
    ),
    title = "Global TB Incidence (per 100,000) Over Time"
  )
```

```{r}
line_plot <- plot_ly(
  global_totals,
  x = ~year,
  y = ~total_inc,
  type = "scatter",
  mode = "lines+markers",
  name = "Global total incident cases",
  hovertemplate = "Year: %{x}<br>Total cases: %{y}<extra></extra>"
) %>%
  layout(
    yaxis = list(title = "Global Incident TB Cases"),
    xaxis = list(title = "Year")
  )


# ---- COMBINE INTO SUBPLOT ----
final_dashboard <- subplot(
  map_plot,
  line_plot,
  nrows = 2,
  heights = c(0.7, 0.3),
  titleY = TRUE
)
```
```{r}
# ---- PREP ----
tb_clean <- tb %>%
  filter(!is.na(e_inc_100k)) %>%
  mutate(year = as.integer(year))

global_totals <- tb_clean %>%
  group_by(year) %>%
  summarise(total_inc = sum(e_inc_num, na.rm = TRUE))


# ---- MAP (no warnings) ----
map_plot <- tb_clean %>%
  plot_geo() %>%
  add_trace(
    type = "choropleth",
    locations = ~iso3,
    locationmode = "ISO-3",
    z = ~e_inc_num,
    animation_frame = ~year,
    colorscale = "Viridis",
    hoverinfo = "text",
    text = ~paste0(
      country, "<br>",
      "Incidence: ", round(e_inc_100k, 1), " per 100k"
    ),
    marker = list(line = list(width = 0))
  ) %>%
  layout(
    title = "Global TB Incidence (per 100,000) Over Time",
    geo = list(
      projection = list(type = "natural earth"),
      showcountries = TRUE,
      showcoastlines = FALSE,
      showframe = FALSE
    )
  )


# ---- LINE PLOT ----
line_plot <- plot_ly(
  global_totals,
  x = ~year,
  y = ~total_inc,
  type = "scatter",
  mode = "lines+markers",
  name = "Global TB Cases"
) %>%
  layout(
    xaxis = list(title = "Year"),
    yaxis = list(title = "Total Incident TB Cases")
  )


# ---- COMBINE ----
final_dashboard <- subplot(
  map_plot,
  line_plot,
  nrows = 2,
  heights = c(0.7, 0.3),
  titleY = TRUE
)

final_dashboard

```







### About the Data

#### Data Dictionary

#### `who_tb_data.csv`

| variable | class | description |
|:---|:---|:---|
| country | character | Country or territory name |
| g_whoregion | character | WHO region |
| iso_numeric | integer | ISO numeric country/territory code |
| iso2 | character | ISO 2-character country/territory code. Note that Namibia's code ("'NA'") includes single quotes to avoid being encoded as missing |
| iso3 | character | ISO 3-character country/territory code |
| year | integer | Year of observation |
| c_cdr | double | Case detection rate (all forms) [also known as TB treatment coverage], percent |
| c_newinc_100k | double | Case notification rate, which is the total of new and relapse cases and cases with unknown previous TB treatment history per 100 000 population (calculated) |
| cfr | double | Estimated TB case fatality ratio |
| e_inc_100k | double | Estimated incidence (all forms) per 100 000 population |
| e_inc_num | integer | Estimated number of incident cases (all forms) |
| e_mort_100k | double | Estimated mortality of TB cases (all forms) per 100 000 population |
| e_mort_exc_tbhiv_100k | double | Estimated mortality of TB cases (all forms, excluding HIV) per 100 000 population |
| e_mort_exc_tbhiv_num | integer | Estimated number of deaths from TB (all forms, excluding HIV) |
| e_mort_num | integer | Estimated number of deaths from TB (all forms) |
| e_mort_tbhiv_100k | double | Estimated mortality of TB cases who are HIV-positive, per 100 000 population |
| e_mort_tbhiv_num | integer | Estimated number of deaths from TB in people who are HIV-positive |
| e_pop_num | integer | Estimated total population number |